### 2009-05-21 paper.tex
comment for revision
small edits

This paper focuses on model construction and inference algorithms
rather than the heuristics and constraints which will be necessary to make these
algorithms fast enough for analysis of many biological datasets.
As discussed below, the complexity of our inference algorithms is
prohibitively high for many problems of interest.
However, this complexity can be significantly
reduced by incorporating outside knowledge.  For example, if we know
the consensus structure of several sequences or their individual
structures, then we can constrain our algorithms accordingly.
Similarly, we might consider only ancestral structures which are
compatible with a curated multiple sequence alignment.  Such constraints are
commonly used by programs for RNA sequence analysis \cite{RivasEddy2001,Holmes2005}.

@@

The stochastic state machine which generates parse trees from the marginal distribution $P(X)$ is a {\bf singlet transducer}.
The machine that absorbs parse trees $X$ and generates modified parse trees $Y$ from the conditional distribution $P(Y | X, T)$ is a {\bf branch transducer}.
Singlet transducers only emit symbols to their output sequence;
branch transducers, in contrast, can both emit symbols to their output
and absorb symbols from their input sequence.

Much as the states of Pair HMMs with no null states, other than the $\Sstart$
and $\Send$ states, can be classified as $\Sinsert$ or $\Smatch$,
transducers can have states of type $\Sstart$, $\Send$,
$\Swait$, $\Sinsert$ and $\Smatch$.
The first three state types, $\Sstart$, $\Send$ and $\Swait$, are null:
They do not emit or absorb any symbols and are required solely for
``organizational'' purposes (see following section).
Two types of states can emit and/or absorb symbols, $\Sinsert$ and $\Smatch$.
An $\Sinsert$ state emits a symbol to the output without absorbing anything.
A $\Smatch$ state absorbs a symbol on the input and either emits
the same symbol to the output, substitutes a different output symbol,
or emits no output symbol at all, the last corresponding to a deletion.

@@

The ensemble model contains empty bifurcations {\em and} empty paths.
An {\bf empty bifurcation} occurs when a child branch of a bifurcation state
transitions to the $\Send$ state without emitting any symbols and can
be removed from the model by creating an effective direct transition
encapsulating the empty bifurcation.  For
example, we can create an effective direct transition $N_1 \rightarrow
N_3$ between null states $N_1$ and $N_3$ in place of the empty
bifurcation $N_1 \rightarrow B \rightarrow (N_2 \,\, N_3) \rightarrow
(End \,\, N_3)$, where $B$ is a bifurcation state with children $(N_2
\,\, N_3)$. Bifurcation states are the most computationally-costly
part of our models, so it is important to eliminate as many as
possible without reducing model expressiveness.

Similarly, our model construction procedure generically produces many
{\bf empty paths} which do not emit any terminal symbols.
Most common are paths of ``windback'' null transitions via non-emitting placeholder states.
If $\Semit$ states $E_1$ and $E_2$ are connected in the state
graph by $\Snull$ states $N_1$ and $N_2$, then the transition $E_1
\rightarrow N_1 \rightarrow N_2 \rightarrow E_2$ with  probability
$P(E_1 \rightarrow N_1) \cdot P(N_1 \rightarrow N_2) \cdot P(N_2
\rightarrow E_2)$ can be replaced with a transition $E_1 \rightarrow
E_2$ with an identical probability.

@@

We refer to empty paths and empty bifurcations as {\bf null cycles}.
They occur (in simpler form) in Hidden Markov Models (which are special cases of SCFGs)
and independent-sites models (which can be viewed as special cases of HMMs).
In all such models, null cycles represent histories that are valid
according to the model but cannot be resolved by direct observation.
Such null events can be real (e.g., ancestral residues that have been deleted in all extant lineages)
or they can be artefactual (e.g., transitions between placeholder null states of an HMM).
Often, one can account for the likelihood of such events by exact marginalization.
For example, eliminating missing ``empty'' columns from an independent-sites model of sequence involves a straightforward application of a correction factor
to account for the proportion of columns that are unobserved \cite{RivasEddy2008}.
The slightly more complicated situation of missing emissions in a HMM can be dealt with by
summing over all null paths, yielding a geometric series that is solvable by matrix inversion
\cite{Holmes2003,Holmes2007,Lunter2007}; the ``wing retraction'' algorithms used in HMMER can be viewed as a special case of matrix inversion by iterated elimination \cite{hmmer}.

Algorithms for exact elimination of HMM null cycles by matrix inversion are implemented
by programs such as PhyloComposer \cite{Holmes2003,Holmes2007} and
HMMoC \cite{Lunter2007}. These algorithms effectively replace the HMM
with an equivalent HMM that contains no null cycles but generates the
same probability distribution over sequences.
However, these solutions do not easily generalize to SCFGs, due to the
possible presence of null bifurcations.

@@

We used our model-construction algorithm to build the grammar
corresponding to the TKFST model acting on a star phylogeny with three
(extant) leaf sequences and a single (unobserved) ancestral
sequence. We chose this phylogeny for two reasons: (1) it is the
simplest extension of the well-studied, standard two-sequence (Pair
SCFG) model and (2) algorithms on a phylogeny with three leaves
should be sufficient for ergodic sampling of reconstructions on any larger phylogeny,
using a MCMC kernel similar to that of \cite{JensenHein2002}.

@@

Guided by our experience with the \emph{nanos} 3' TCE and tRNA, where
the reconstruction problem was made easy by the presence of a close
outgroup, we conducted a simulation study to investigate the
dependence of reconstruction accuracy on outgroup branch length.
We simulated the evolution of RNAs under the TKFST model along
3-taxon phylogenies, where we kept the branch lengths of the two
sibling species constant and varied the branch length of the
outgroup.  We chose the parameters of the TKFST model to
generate ``interesting'' alignments ($\sim 0.3$ substructure indels per
alignment) and considered only alignments with at least two ancestral
stems, sequences of length $\in [30, 70]$ bp, loops of length $\in [3,
10]$ bp and stems of lengths $\in [1, 20]$ bp.

@@

While these results correspond to a best-case scenario in which we
have a perfectly-parametrized evolutionary model, recent results
demonstrating that inference of sequence alignment is surprisingly
robust to mis-parametrized probabilistic models \cite{Lunter2007}
suggest that the same may be true for probabilistic inference of
ancestral structures.

>>>>

This paper focuses on model construction and inference algorithms
rather than the heuristics and constraints which will be necessary to make these
algorithms fast enough for analysis of many biological datasets.
As discussed below, the complexity of general inference algorithms is
prohibitively high for many problems of interest.
However, this complexity can be significantly
reduced by incorporating outside knowledge.  For example, if we know
the consensus structure of several sequences or their individual
structures, then we can constrain our algorithms accordingly.
Similarly, we might consider only ancestral structures which are
compatible with a given multiple sequence alignment, or a relatively small set of candidate alignments (as in the ORTHEUS program \cite{PatenHolmesBirney2008}). 
Such constraints are
commonly used by programs for SCFG-based RNA sequence analysis such as QRNA \cite{RivasEddy2001}, Stemloc \cite{Holmes2005} and CONSAN \cite{DowellEddy2006}.
Alignment and structural constraints can be combined \cite{Holmes2005}.

@@

Consider the stochastic grammar which generates parse trees from the marginal distribution $P(X)$.
It is occasionally convenient to represent this grammar as a transducer whose input is constrained to be null,
i.e. a machine that accepts a dummy (empty) input sequence, and outputs sequence $X$.
We refer to this as the {\bf singlet transducer}.
In contrast, the more general type of transducer that absorbs parse trees $X$ and generates modified parse trees $Y$ from the conditional distribution $P(Y | X, T)$ is a {\bf branch transducer}.
By definition, singlet transducers only emit symbols to their output sequence, and use a restricted set of state types.
Branch transducers, in contrast, can both emit symbols to their outputs and absorb symbols from their inputs, and so use the full range of state types.

Transducers can have states of type $\Sstart$, $\Send$,
$\Swait$, $\Sinsert$ and $\Smatch$.
The first three state types, $\Sstart$, $\Send$ and $\Swait$, are null:
they do not emit or absorb any symbols and are required solely for
``organizational'' purposes (see following section).
Two types of states can emit and/or absorb symbols, $\Sinsert$ and $\Smatch$.
An $\Sinsert$ state emits a symbol to the output without absorbing anything.
A $\Smatch$ state absorbs a symbol on the input and either emits
the same symbol to the output, substitutes a different output symbol,
or emits no output symbol at all, the last corresponding to a deletion.

@@

The ensemble grammar contains many rules that can be applied redundantly,
together or in isolation, to generate subtrees of the parse tree that do
not generate any terminals, and could be replaced with simpler rules.
In fact, this is a general property of SCFGs with null states and/or bifurcations.
In this subsection, we describe how to transform the SCFG so as to eliminate such redundant rules.
We begin by identifying two types of redundant parse-structures: {\bf empty bifurcations} and {\bf empty paths}.

An {\bf empty bifurcation} occurs when a child branch of a bifurcation state
transitions to the $\Send$ state without emitting any symbols and can
be removed from the model by creating an effective direct transition
encapsulating the empty bifurcation.  For
example, we can create an effective direct transition $N_1 \rightarrow
N_3$ between null states $N_1$ and $N_3$ in place of the empty
parse-subtree $N_1 \rightarrow B \rightarrow (N_2 \,\, N_3) \rightarrow
(End \,\, N_3)$, where $B$ is a bifurcation state with children $(N_2
\,\, N_3)$. Bifurcation states are the most computationally-costly
part of our models, so it is important to eliminate as many as
possible without reducing model expressiveness.

In contrast, an {\bf empty path} is defined as any parse-subtree {\em without} bifurcations that does not emit terminal symbols.
Most common are paths of ``windback'' null transitions via non-emitting placeholder states.
If $\Semit$ states $E_1$ and $E_2$ are connected in the state
graph via $\Snull$ states $N_1$ and $N_2$, then the path $E_1
\rightarrow N_1 \rightarrow N_2 \rightarrow E_2$ with  probability
$P(E_1 \rightarrow N_1) \cdot P(N_1 \rightarrow N_2) \cdot P(N_2
\rightarrow E_2)$ can be replaced by a single direct transition $E_1 \rightarrow
E_2$ with an identical probability.

@@

Empty paths occur in Hidden Markov Models (which are special cases of SCFGs)
and independent-sites models (which can be viewed as special cases of HMMs).
Conceptually, empty paths can represent histories that are valid
according to the model but cannot be resolved by direct observation.
Such null events can be real (e.g., ancestral residues that have been deleted in all extant lineages)
or they can be artefactual (e.g., transitions between placeholder null states of an HMM).

We refer to empty paths and empty bifurcations collectively as {\bf null cycles}.
Our goal is an algorithmic procedure to account for the likelihood of such cycles by exact marginalization.
For simpler models, solutions to this problem are published.
Eliminating missing (empty) columns from an independent-sites model of sequence involves a straightforward application of a correction factor
to account for the proportion of columns that are unobserved \cite{RivasEddy2008}.
The slightly more complicated situation of missing emissions in a HMM can be dealt with by
summing over all empty paths, yielding a geometric series that is solvable by matrix inversion
\cite{Holmes2003,Holmes2007,Lunter2007}.

Algorithms for exact elimination of empty paths from HMMs by matrix inversion are implemented
by programs such as PhyloComposer \cite{Holmes2003,Holmes2007} and
HMMoC \cite{Lunter2007}. These algorithms effectively replace the HMM
with an equivalent HMM that contains no null cycles but generates the
same probability distribution over sequences.
The ``wing retraction'' algorithms used in the repetitively structured profile HMMs of HMMER \cite{hmmer} appear similar in operation.
However, these solutions do not easily generalize to SCFGs, which contain bifurcations
(and, therefore, may generate empty bifurcations as well as empty paths).

@@

We used our model-construction algorithm to build the grammar
corresponding to the TKFST model acting on a star phylogeny with three
(extant) leaf sequences and a single (unobserved) ancestral
sequence. We chose this phylogeny for two reasons: (1) it is the
simplest extension of the well-studied, standard two-sequence (Pair
SCFG) model and (2) algorithms on a phylogeny with three leaves
should be sufficient for ergodic sampling of reconstructions on any larger phylogeny,
using e.g. a Gibbs-sampling MCMC kernel \cite{JensenHein2002}
or a progressive suboptimal-alignment sampling heuristic \cite{PatenHolmesBirney2008}.

@@

Guided by our experience with the \emph{nanos} 3' TCE and tRNA, where
the reconstruction problem was made easy by the presence of a close
outgroup, we conducted a simulation study to investigate the
dependence of reconstruction accuracy on outgroup branch length.
We simulated the evolution of RNAs under the TKFST model along
3-taxon phylogenies (with one internal node), where we kept the branch lengths of two
sibling species constant and varied the branch length of the
outgroup.  We chose the parameters of the TKFST model to
generate ``interesting'' alignments ($\sim 0.3$ substructure indels per
alignment) and considered only alignments with at least two ancestral
stems, sequences of length $\in [30, 70]$ bp, loops of length $\in [3,
10]$ bp and stems of lengths $\in [1, 20]$ bp.
We then used \indiegram\ to reconstruct maximum likelihood reconstructions of the ancestral structure and the four-way multiple alignment (ancestor + three descendants).

@@

While these results correspond to a best-case scenario in which we
have a perfectly-parametrized evolutionary model, recent results
demonstrating that inference of sequence alignment is
robust to mis-parametrized probabilistic models \cite{Lunter2007}
suggest that the same may be true for probabilistic inference of
ancestral structures.
